---
title: "1. Coding in R (9 points)"
author: "Miguel Naranjo, 97943"
date: "2023-10-08"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
```

# Card Game Simulator

Implementation of the card game simulator in R (Fig. 1). Implementation MUST be clearly explained in a PDF document created by using R markdown.

1.  **Initialize the Deck**:

    -   We start by initializing a deck of cards. In this script, we represent the deck as a vector containing cards with values 2 to 11 (representing number cards and aces) and 10 (representing face cards like Jack, Queen, King). The deck is shuffled to ensure randomness.

        ```{r}
        # Create a vector of card ranks and suits
        ranks <- c("2", "3", "4", "5", "6", "7", "8", "9", "10", "Jack", "Queen", "King", "Ace")
        suits <- c("Hearts", "Diamonds", "Clubs", "Spades")

        # Create an empty data frame to store the cards
        deck <- data.frame(Rank = character(0), Suit = character(0))

        # Loop through ranks and suits to generate all cards
        for (suit in suits) {
          for (rank in ranks) {
            card <- data.frame(Rank = rank, Suit = suit)
            deck <- rbind(deck, card)
          }
        }

        # Write the deck of cards to a CSV file
        write.csv(deck, "poker_deck.csv", row.names = FALSE)

        ```

2.  **Initialize Player and Dealer Hands**:

    -   We create empty vectors for the player's and dealer's hands to hold the cards they receive.

        ```{r}
        # Function to read a CSV file and add a "Value" column to the data frame
        read_deck <- function(file_name) {
          # Read the CSV file into a data frame
          deck <- read.csv(file_name, header = TRUE)
          
          # Define card values
          card_values <- c(
            "2" = 2, "3" = 3, "4" = 4, "5" = 5, "6" = 6, "7" = 7, "8" = 8, "9" = 9, "10" = 10,
            "Jack" = 11, "Queen" = 12, "King" = 13, "Ace" = 14
          )
          
          # Add a "Value" column based on card ranks
          deck$Value <- card_values[deck$Rank]
          
          return(deck)
        }

        # Example usage:
        file_name <- "poker_deck.csv"  # Replace with the actual CSV file name
        deck_data <- read_deck(file_name)

        # Print the resulting data frame
        print(deck_data)

        ```

3.  **Game Loop**:

    -   The game runs in a continuous loop (**`while (TRUE)`**) until the player decides to stop.

        ```{r}
        # Function to shuffle a deck of cards represented as a data frame
        shuffle_deck <- function(deck) {
          # Use the sample function to shuffle the rows of the data frame
          shuffled_deck <- deck[sample(nrow(deck)), ]
          rownames(shuffled_deck) <- NULL  # Reset row names
          
          return(shuffled_deck)
        }

        # Example usage:
        # Assuming you have a deck data frame, you can shuffle it like this:
        shuffled_deck <- shuffle_deck(deck_data)

        # Print the shuffled deck
        print(shuffled_deck)

        ```

4.  **Player's Turn**:

    -   During the player's turn, the script displays the player's hand and one of the dealer's cards.

    -   The player is asked to choose between "hit" (take another card) or "stand" (stop taking cards).

    -   If the player chooses to hit, a card is dealt from the deck and added to the player's hand. The hand value is calculated.

    -   If the player chooses to stand, the player's turn ends.

        ```{r}
        # Function to deal n cards from a shuffled deck, ensuring no card is dealt twice
        deal_cards <- function(deck, n) {
          # Initialize an empty list to keep track of dealt cards
          dealt_cards <- list()
          
          # Initialize an empty data frame to store the dealt cards
          dealt_deck <- data.frame(Rank = character(0), Suit = character(0), Value = numeric(0))
          
          # Check if there are enough cards left to deal
          if (n > nrow(deck)) {
            stop("Not enough cards left in the deck to deal ", n, " cards.")
          }
          
          for (i in 1:n) {
            # Check if all cards have been dealt
            if (nrow(deck) == 0) {
              warning("All cards have been dealt.")
              break
            }
            
            # Randomly select a card from the remaining deck
            random_index <- sample(nrow(deck), 1)
            dealt_card <- deck[random_index, ]
            
            # Remove the dealt card from the deck
            deck <- deck[-random_index, , drop = FALSE]
            
            # Add the dealt card to the dealt cards list and data frame
            dealt_cards[[i]] <- dealt_card
            dealt_deck <- rbind(dealt_deck, dealt_card)
          }
          
          return(dealt_deck)
        }

        # Example usage:
        # Assuming you have a shuffled deck represented by the 'shuffled_deck' data frame, you can deal 5 cards like this:
        dealt_cards <- deal_cards(shuffled_deck, n = 5)

        # Print the dealt cards
        print(dealt_cards)

        ```

5.  **Check for Player Bust or Blackjack**:

    -   After each player's action, we check if the player's hand value exceeds 21 (bust) or equals 21 (blackjack). In case of a bust or blackjack, the game round ends.

        ```{r}
        # Function to calculate the total value of a hand
        calculate_hand_value <- function(hand) {
          hand_value <- sum(hand)
          
          # Check for aces and adjust for soft hands
          if (any(hand == 11) && hand_value > 21) {
            num_aces <- sum(hand == 11)
            while (num_aces > 0 && hand_value > 21) {
              hand_value <- hand_value - 10
              num_aces <- num_aces - 1
            }
          }
          
          return(hand_value)
        }

        # Function to deal a new card
        deal_card <- function(deck, hand) {
          card <- deck[1]
          deck <- deck[-1]
          hand <- c(hand, card)
          return(list(deck, hand))
        }

        # Initialize the deck
        deck <- c(2, 3, 4, 5, 6, 7, 8, 9, 10, 10, 10, 10, 11)
        deck <- rep(deck, 4)

        # Shuffle the deck
        deck <- sample(deck)

        # Initialize player and dealer hands
        player_hand <- numeric(0)
        dealer_hand <- numeric(0)

        # Deal the initial two cards to the player and dealer
        deck_and_player <- deal_card(deck, player_hand)
        deck <- deck_and_player[[1]]
        player_hand <- deck_and_player[[2]]

        deck_and_dealer <- deal_card(deck, dealer_hand)
        deck <- deck_and_dealer[[1]]
        dealer_hand <- deck_and_dealer[[2]]

        # Game loop
        while (TRUE) {
          # Calculate hand values
          player_value <- calculate_hand_value(player_hand)
          dealer_value <- calculate_hand_value(dealer_hand)
          
          # Display hands
          cat("Your hand:", paste(player_hand, collapse = ", "), "\n")
          cat("Dealer's hand:", paste(dealer_hand[1], "??"), "\n")
          
          # Check for blackjack
          if (player_value == 21) {
            cat("Blackjack! You win!\n")
            break
          }
          
          # Player's turn
          action <- readline(prompt = "Do you want to hit or stand? (h/s): ")
          
          if (action == "h") {
            deck_and_player <- deal_card(deck, player_hand)
            deck <- deck_and_player[[1]]
            player_hand <- deck_and_player[[2]]
          } else if (action == "s") {
            break
          }
          
          # Check for player bust
          if (player_value > 21) {
            cat("Bust! You lose!\n")
            break
          }
          
          # Dealer's turn
          while (dealer_value < 17) {
            deck_and_dealer <- deal_card(deck, dealer_hand)
            deck <- deck_and_dealer[[1]]
            dealer_hand <- deck_and_dealer[[2]]
            dealer_value <- calculate_hand_value(dealer_hand)
          }
          
          # Check for dealer bust or compare hands
          if (dealer_value > 21) {
            cat("Dealer busts! You win!\n")
          } else if (dealer_value == player_value) {
            cat("It's a tie! Push.\n")
          } else if (dealer_value > player_value) {
            cat("Dealer wins!\n")
          } else {
            cat("You win!\n")
          }
          
          break
        }

        cat("Thanks for playing!\n")

        ```

6.  **Dealer's Turn**:

    -   The dealer's turn begins. The dealer will draw cards until their hand value reaches 17 or higher.

    -   Cards are dealt to the dealer from the deck, and the hand value is calculated after each card.

        ```{r}
        # Function to calculate the total value of a hand
        calculate_hand_value <- function(hand) {
          hand_value <- sum(hand)
          
          # Check for aces and adjust for soft hands
          if (any(hand == 11) && hand_value > 21) {
            num_aces <- sum(hand == 11)
            while (num_aces > 0 && hand_value > 21) {
              hand_value <- hand_value - 10
              num_aces <- num_aces - 1
            }
          }
          
          return(hand_value)
        }

        # Function to deal a new card
        deal_card <- function(deck, hand) {
          card <- deck[1]
          deck <- deck[-1]
          hand <- c(hand, card)
          return(list(deck, hand))
        }

        # Initialize the deck
        deck <- c(2, 3, 4, 5, 6, 7, 8, 9, 10, 10, 10, 10, 11)
        deck <- rep(deck, 4)

        # Shuffle the deck
        deck <- sample(deck)

        # Initialize player and dealer hands
        player_hand <- numeric(0)
        dealer_hand <- numeric(0)

        # Initialize variables to track wins and losses
        wins <- 0
        losses <- 0

        # Game loop
        while (TRUE) {
          # Calculate hand values
          player_value <- calculate_hand_value(player_hand)
          dealer_value <- calculate_hand_value(dealer_hand)
          
          # Display hands
          cat("Your hand:", paste(player_hand, collapse = ", "), "\n")
          cat("Dealer's hand:", paste(dealer_hand[1], "??"), "\n")
          
          # Check for blackjack
          if (player_value == 21) {
            cat("Blackjack! You win!\n")
            wins <- wins + 1
            break
          }
          
          # Player's turn
          action <- readline(prompt = "Do you want to hit or stand? (h/s): ")
          
          if (action == "h") {
            deck_and_player <- deal_card(deck, player_hand)
            deck <- deck_and_player[[1]]
            player_hand <- deck_and_player[[2]]
          } else if (action == "s") {
            break
          }
          
          # Check for player bust
          if (player_value > 21) {
            cat("Bust! You lose!\n")
            losses <- losses + 1
            break
          }
          
          # Dealer's turn
          while (dealer_value < 17) {
            deck_and_dealer <- deal_card(deck, dealer_hand)
            deck <- deck_and_dealer[[1]]
            dealer_hand <- deck_and_dealer[[2]]
            dealer_value <- calculate_hand_value(dealer_hand)
          }
          
          # Check for dealer bust or compare hands
          if (dealer_value > 21) {
            cat("Dealer busts! You win!\n")
            wins <- wins + 1
          } else if (dealer_value == player_value) {
            cat("It's a tie! Push.\n")
          } else if (dealer_value > player_value) {
            cat("Dealer wins!\n")
            losses <- losses + 1
          } else {
            cat("You win!\n")
            wins <- wins + 1
          }
          
          # Ask the user if they want to continue playing
          continue_playing <- readline(prompt = "Do you want to continue playing? (Y/N): ")
          if (toupper(continue_playing) != "Y") {
            break
          }
          
          # Reset hands for the next round
          player_hand <- numeric(0)
          dealer_hand <- numeric(0)
        }

        # Save the results to a CSV file
        results <- data.frame(Wins = wins, Losses = losses)
        write.csv(results, "blackjack_results.csv", row.names = FALSE)

        cat("Thanks for playing!\n")

        ```

7.  **Check for Dealer Bust or Compare Hands**:

    -   After the dealer's turn, we check if the dealer's hand value exceeds 21 (bust).

    -   If the dealer busts, the player wins.

    -   If not, we compare the final hand values of the player and the dealer to determine the winner.

8.  **Recording Wins and Losses**:

    -   We keep track of the number of wins and losses for the player throughout the game using the **`wins`** and **`losses`** variables.

9.  **Ask to Continue Playing**:

    -   After each round, the player is asked if they want to continue playing. If the player enters "Y" (yes), a new round begins. If the player enters "N" (no), the game loop exits.

10. **Saving Results**:

    -   At the end of the game, the script creates a data frame (**`results`**) to store the number of wins and losses.

    -   It uses the **`write.csv`** function to save these results in a CSV file named "blackjack_results.csv."

11. **Exit the Game**:

    -   The script thanks the player for playing and exits.
